---
title: "Module05 homework, Q3"
author: "Steve Simon"
output:
  pdf_document: default
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, 
    encoding=encoding,
    output_dir = "../results", 
    output_format = "all") })  
---

Dates: This file was created on 2020-07-14 and last modified on `r Sys.Date()`.

Purpose: Answer M05-Q03. Use the ehr database and write a SQL query that will list all the table names.

Note: this solution uses R and SQLite. An alternate solution using SAS and Oracle is also available.

Note: Some of the names used in this code are arbitrary and you can choose whatever names you want. To emphasize which names can be modified at your discretion, I am using names of famous statisticians.

The statistician being honored in this code is [George Box](https://en.wikipedia.org/wiki/George_E._P._Box).

Here is the approach using SQL commands.

```{r george_box1}
library(sqldf)
box <- dbConnect(SQLite(),
  dbname="../data/ehr.sqlite")
george <- dbGetQuery(conn=box,  "
    select tbl_name from sqlite_master
")

george
dbDisconnect(conn=box)
```

Here is an alternative approach using an R function. You don't have to do your homework both ways; I'm just showing different methods.

```{r george_box2}
library(sqldf)
box <- dbConnect(SQLite(),
  dbname="../data/ehr.sqlite")
dbListTables(conn=box)
dbDisconnect(conn=box)
```
