---
title: "Homework06a"
author: "Steve Simon"
output:
  pdf_document: default
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, 
    encoding=encoding,
    output_dir = "../results", 
    output_format = "all") })  
---

This file was created on 2020-07-06 and last modified on `r Sys.Date()`.

Note: this solution uses R and SQLite. An alternate solution using SAS and Oracle is also available.

  For your homework, use the hospital database that I mentioned briefly
  earlier in this lecture.

  1. Verify that the hospital id code (HOSP_ID) has no missing values.
  2. There are only two fields in the database that have null values.
  Get a count of the number of missing values for the indicator for
  teaching hospital (TEACHING_IND).
  3. There is only one hospital where the number of beds (BED_SIZE) is
  unknown. Find the id of that hospital.
  4. Combine your results into a single PDF file and submit it.;

Note: Some of the names used in this code are arbitrary and you can choose whatever names you want. To emphasize which names can be modified at your discretion, I am using names of famous statisticians.

The statistician being honored in this code is [David Blackwell](https://en.wikipedia.org/wiki/David_Blackwell).

```{r david_blackwell1}
library(sqldf)
blackwell <- dbConnect(SQLite(),
  dbname="../data/hospital.sqlite")
david1 <- dbGetQuery(conn=blackwell,  "
    select
      count(*) as number_missing_ids
    from hospital
    where HOSP_ID is null
")

david1
dbDisconnect(conn=blackwell)
```

```{r david_blackwell2}
library(sqldf)
blackwell <- dbConnect(SQLite(),
  dbname="../data/hospital.sqlite")
david2 <- dbGetQuery(conn=blackwell,  "
    select
      count(*) as number_missing_teaching_ind
      from hospital
      where TEACHING_IND is null
")

david2
dbDisconnect(conn=blackwell)
```

```{r david_blackwell3}
library(sqldf)
blackwell <- dbConnect(SQLite(),
  dbname="../data/hospital.sqlite")
david3 <- dbGetQuery(conn=blackwell,  "
    select
      HOSP_ID
      from hospital
      where BED_SIZE is null
")

david3
dbDisconnect(conn=blackwell)
```
