---
title: "Homework09a"
author: "Steve Simon"
output:
  pdf_document: default
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, 
    encoding=encoding,
    output_dir = "../results", 
    output_format = "all") })  
---

This file was created on 2020-07-12 and last modified on `r Sys.Date()`.

Note: this solution uses R and SQLite. An alternate solution using SAS and Oracle is also available.

Graded: 3 points, one point deduction if submitted after deadline

–Put your code and the output in a single PDF file

–Use Encounter Table

–Use case expression to classify age <= 40 as ‘Group 1’, and age > 40 as ‘Group 2’ 

 

–Use hospital table

–Use coalesce function to return -1 for null values of teaching_ind in hospital table where census_reg = ‘West’

Note: Some of the names used in this code are arbitrary and you can choose whatever names you want. To emphasize which names can be modified at your discretion, I am using names of famous statisticians.

The statistician being honored in this code is [Barbara A. Bailar](https://en.wikipedia.org/wiki/Barbara_A._Bailar).

```{r barbara_bailar1}
library(sqldf)
bailar <- dbConnect(SQLite(),
  dbname="../data/hospital.sqlite")
barbara1 <- dbGetQuery(conn=bailar,  "
  select 
    teaching_ind,
    coalesce(teaching_ind, -1) as imputed_value
  from hospital
  where census_reg='West'
")

barbara1
dbDisconnect(conn=bailar)
```

```{r barbara_bailar2}
library(sqldf)
bailar <- dbConnect(SQLite(),
  dbname="../data/encounter_db.sqlite")
barbara2 <- dbGetQuery(conn=bailar,  "
  select 
    age,
    case 
      when age <= 40
        then 'Group 1'
        else 'Group 2'
      end as age_group
  from encounter
")

barbara2
dbDisconnect(conn=bailar)
```

