---
title: "Homework10a"
author: "Steve Simon"
output:
  pdf_document: default
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, 
    encoding=encoding,
    output_dir = "../results", 
    output_format = "all") })  
---

This file was created on 2020-07-14 and last modified on `r Sys.Date()`.

Note: this solution uses R and SQLite. An alternate solution using SAS and Oracle is also available.

Note: Some of the names used in this code are arbitrary and you can choose whatever names you want. To emphasize which names can be modified at your discretion, I am using names of famous statisticians.

The statistician being honored in this code is [George Box](https://en.wikipedia.org/wiki/George_E._P._Box).

1. Pick a database (any database)
  +	Use one of the approaches shown above to list all the table names.

Here is the approach using SQL commands.

```{r george_box1}
library(sqldf)
box <- dbConnect(SQLite(),
  dbname="../data/hospital.sqlite")
george1 <- dbGetQuery(conn=box,  "
    select tbl_name from sqlite_master
")

george1
dbDisconnect(conn=box)
```

Here is an alternative approach using an R function. You don't have to do your homework both ways; I'm just showing different methods.

```{r george_box2}
library(sqldf)
box <- dbConnect(SQLite(),
  dbname="../data/hospital.sqlite")
dbListTables(conn=box)
dbDisconnect(conn=box)
```

2. Pick a table (any table)
  +	Use one of the approaches shown above to list all the field names.

Here is the approach using SQL commands.

```{r george_box3}
library(sqldf)
box <- dbConnect(SQLite(),
  dbname="../data/hospital.sqlite")
george3 <- dbGetQuery(conn=box,  "
    pragma table_info(hospital)
")

george3
dbDisconnect(conn=box)
```

Here is an alternative approach using an R function. You don't have to do your homework both ways; I'm just showing different methods.

```{r george_box4}
library(sqldf)
box <- dbConnect(SQLite(),
  dbname="../data/hospital.sqlite")
dbListFields(conn=box, name="hospital")
dbDisconnect(conn=box)
```

3. Do an Internet search on a database other than Oracle and SQLite.
  +	Document how you get a list of all the table names in that database.

I ran a search on the phrase "mysql list all table names" and found a web page

https://stackoverflow.com/questions/8334493/get-table-names-using-select-statement-in-mysql

This talks in a bit more detail than I needed, but the basic answer is to use the 
SQL statements

```{}
  SELECT table_name FROM information_schema.tables
  WHERE table_schema = 'your_database_name';
```

