---
title: "Homework13a"
author: "Steve Simon"
output:
  pdf_document: default
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, 
    encoding=encoding,
    output_dir = "../results", 
    output_format = "all") })  
---

This file was created on 2020-07-24 and last modified on `r Sys.Date()`.

Note: this solution uses R and SQLite. An alternate solution using SAS and Oracle is also available.

Note: Some of the names used in this code are arbitrary and you can choose whatever names you want. To emphasize which names can be modified at your discretion, I am using names of famous statisticians.

The statistician being honored in this code is [Hirotugu Akaike](https://en.wikipedia.org/wiki/Hirotugu_Akaike).

1. Do an inner join of acupuncture_baseline_results and acupuncture_one_year_results. Display the first ten rows of data only.

```{r hirotugu_akaike1}
library(sqldf)
akaike <- dbConnect(SQLite(),
  dbname="../data/melange.sqlite")
hirotugu1 <- dbGetQuery(conn=akaike,  "
    select b.id, b.pk1, o.pk5
      from acupuncture_baseline_results as b
      join acupuncture_one_year_results as o
      on b.id=o.id
    limit 10
")

hirotugu1
dbDisconnect(conn=akaike)
```

2. Do a left join of acupuncture_baseline_results and acupuncture_one_year_results and display the fields id, pk1, and pk5. Display the first ten rows of data only. Explain why these results are different from the results of the previous question.

```{r hirotugu_akaike2}
library(sqldf)
akaike <- dbConnect(SQLite(),
  dbname="../data/melange.sqlite")
hirotugu2 <- dbGetQuery(conn=akaike,  "
    select b.id, b.pk1, o.pk5
      from acupuncture_baseline_results as b
      left join acupuncture_one_year_results as o
      on b.id=o.id
    limit 10
")

hirotugu2
dbDisconnect(conn=akaike)
```

There are several patients in the baseline table that are not in the one year table. These patients are excluded from the inner join, but not the left join.


3. Count the number of records after an inner join of acupuncture_baseline_results and acupuncture_one_year_results. Count the number of records after a left join of acupuncture_baseline_results and acupuncture_one_year_results.

```{r hirotugu_akaike3}
library(sqldf)
akaike <- dbConnect(SQLite(),
  dbname="../data/melange.sqlite")
hirotugu3a <- dbGetQuery(conn=akaike,  "
    select count(*) as n
      from acupuncture_baseline_results as b
      join acupuncture_one_year_results as o
      on b.id=o.id
")
hirotugu3a
hirotugu3b <- dbGetQuery(conn=akaike,  "
    select count(*) as n
      from acupuncture_baseline_results as b
      left join acupuncture_one_year_results as o
      on b.id=o.id
")
hirotugu3b
dbDisconnect(conn=akaike)
```

4. Compute the average pk score at baseline, the average score at one year, and the average change score.

```{r hirotugu_akaike4}
library(sqldf)
akaike <- dbConnect(SQLite(),
  dbname="../data/melange.sqlite")
hirotugu4 <- dbGetQuery(conn=akaike,  "
    select 
      avg(b.pk1) as pk1_avg,
      avg(o.pk5) as pk5_avg,
      avg(b.pk1)-avg(o.pk5) as change_score
      from acupuncture_baseline_results as b
      join acupuncture_one_year_results as o
      on b.id=o.id
")

hirotugu4
dbDisconnect(conn=akaike)
```

5. Display all the pk1 values for patients 64 and older.

```{r hirotugu_akaike5}
library(sqldf)
akaike <- dbConnect(SQLite(),
  dbname="../data/melange.sqlite")
hirotugu5 <- dbGetQuery(conn=akaike,  "
    select 
      d.id, d.age, b.pk1
      from acupuncture_demographics as d
      inner join acupuncture_baseline_results as b
      on d.id=b.id
      where d.age >= 64
")

hirotugu5
dbDisconnect(conn=akaike)
```

6. There are 100 patients with baseline values but no values at one year. Print the ids of the first ten of these patients.

```{r hirotugu_akaike6}
library(sqldf)
akaike <- dbConnect(SQLite(),
  dbname="../data/melange.sqlite")
hirotugu6 <- dbGetQuery(conn=akaike,  "
    select 
      b.id as unmatched_ids
      from acupuncture_baseline_results as b
      left join acupuncture_one_year_results as o
      on b.id=o.id
    where o.id is null
    limit 10
")

hirotugu6
dbDisconnect(conn=akaike)
```

