---
title: "Simple SQL examples within R: input files"
author: "Steve Simon"
date: "September 12, 2017"
output: html_document
---

The emphasis in this class will be on getting information OUT of a SQL database, not getting information INTO a database. But I am showing the code here just for the sake of completeness.

There are several ways to access SQL databases in R. I'm following an example that appears at 

https://www.r-bloggers.com/r-and-sqlite-part-1/

Run this install.packages function once. Run the library function at the top of every program.

```{r install-packages}
# install.packages("sqldf")
library(magrittr)
library(sqldf)
```

The sqldf package allows you to use SQL queries on an existing data frame in addition to running SQL queries on a SQL database.

When you install sqldf, you also end up installing rsqlite and dbi. 

Run the database creation steps once.

The dbConnect command will connect you to an existing database, but if the dbname argument refers to a file that has not yet been created, dbConnect will create a new SQLite database.

The "official" SQL approach would use the CREATE TABLE and INSERT INTO commands, but you can combine both of these into a single function call to the dbWriteTable function.

I had to tweak the original field names slightly. The field case_number was originally called case, but case is a reserved word in sql. The fields fat_b and fat_s were originally called fat.b and fat.s, but the dot is reserved for sql queries involving multiple tables to help resolve ambiguities when some of the field names are the same across multiple tables.

```{r create a new database}
db <- dbConnect(SQLite(), dbname="../data/fat.sqlite")
dbWriteTable(conn=db, name="fat", value="../data/fat.csv",
  overwrite=TRUE, row.names=FALSE, header=TRUE)
dbDisconnect(conn=db)
```

The dbListTables function will list the names of all the tables in your database.

There is also a nice function for reading an entire table into R, dbReadTable, but I would encourage you to access the entire table using the SQL statement "select * from fat".

```{r test access}
library(sqldf)
db <- dbConnect(SQLite(), dbname="../data/fat.sqlite")
dbListTables(conn=db)
fd <- dbReadTable(conn=db, name="fat")
head(fd)
fd <- dbGetQuery(conn=db, "select * from fat")
head(fd)
dbDisconnect(conn=db)
```

Let's put all of this into a single function to simplify things.

```{r function}
add_to_sqlite <- function(rawdata, sqlite_name, table_name) {
  db <- dbConnect(SQLite(), dbname=sqlite_name)
  dbWriteTable(conn=db, name=table_name, value=rawdata,
    overwrite=TRUE, row.names=FALSE, header=TRUE)
  dbDisconnect(conn=db)
  
  db <- dbConnect(SQLite(), dbname=sqlite_name)
  print(dbListTables(conn=db))
  print(dbGetQuery(conn=db, paste0("select * from ", table_name, " limit 5")))
  dbDisconnect(conn=db)
}
```

Here are some additoinal files that I need to read in SQLite.

This is how to read in the Titanic data set.

```{r titanic}
internet_name <- "http://www.statsci.org/data/general/titanic.txt"
csv_name <- "../data/titanic.csv"
sqlite_name <- "../data/titanic_db.sqlite"
table_name <- "titanic_table"

ti <- read.table(file=internet_name, sep="\t", header=TRUE)
write.csv(ti[-1, ], file=csv_name, row.names=FALSE)

add_to_sqlite(ti, sqlite_name, table_name)
```

This is the pulmonary data set. A description of the files is at

http://jse.amstat.org/datasets/fev.txt

```{r pulmonary-data}
internet_name <- "http://www.amstat.org/publications/jse/datasets/fev.dat.txt"
csv_name <- "../data/fev.csv"
sqlite_name <- "../data/fev.sqlite"
table_name <- "fev"

pu <- read.table(file=internet_name)
names(pu) <- c("age", "fev", "ht", "sex", "smoke")
write.csv(pu, file=csv_name)

add_to_sqlite(pu, sqlite_name, table_name)
```

The cigarettes file was downloaded from the Data Description website.

```{r cigarettes}
csv_name <- "../data/cigarettes.csv"
sqlite_name <- "../data/cigarettes_db.sqlite"
table_name <- "cigarettes_table"

cig <- read.csv(file=csv_name, header=TRUE, as.is=TRUE)

add_to_sqlite(cig, sqlite_name, table_name)
```

The airlines file was also found at the Data Description website.

```{r airlines}
csv_name <- "../data/airlines.csv"
sqlite_name <- "../data/airlines_db.sqlite"
table_name <- "airlines_table"

air <- read.csv(file=csv_name, header=TRUE, as.is=TRUE)

add_to_sqlite(air, sqlite_name, table_name)
```

The crawling file was also found at the Data Description website.

```{r crawling}
csv_name <- "../data/crawling.csv"
sqlite_name <- "../data/crawling_db.sqlite"
table_name <- "crawling_table"

crawl <- read.csv(file=csv_name, header=TRUE, as.is=TRUE)

add_to_sqlite(crawl, sqlite_name, table_name)
```

Read additional data from a PDF file.

```{r pdf}
csv_name <- "../data/pdf.csv"
pdf_name <- "../data/connecting.pdf"
sqlite_name <- "hospital"
table_name <- "hospital"

library(pdftools)
txt <- pdf_text("../data/connecting.pdf")
txt[5] %>%
  strsplit("\r\n") %>%
  unlist %>%
  gsub("^\\s+", "", .) %>%
  gsub("\\s+", ",", .) %>%
  writeLines(csv_name)
pdf <- read.csv(csv_name, header=FALSE, skip=2, as.is=TRUE)
names(pdf) <- c("I", "HOSP_ID", "CENSUS_REG", "BED_SIZE", "TEACHING_IND", "RURAL_URBAN", "ACUTE_NONACUTE")

add_to_sqlite(pdf[-1, ], sqlite_name, table_name)
```

```{r save-everything}
save.image(file="../data/input-files.RData")
```