---
title: "Module05 homework, Q4"
author: "Steve Simon"
output:
  pdf_document: default
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, 
    encoding=encoding,
    output_dir = "../results", 
    output_format = "all") })  
---

Dates: This file was created on 2020-07-14 and last modified on `r Sys.Date()`.

Purpose: Answer M05-Q04. Use the melange database and write a SQL query that will list all the field names in the crawling table.

Note: this solution uses R and SQLite. An alternate solution using SAS and Oracle is also available.

Note: Some of the names used in this code are arbitrary and you can choose whatever names you want. To emphasize which names can be modified at your discretion, I am using names of famous statisticians.

The statistician being honored in this code is [George Box](https://en.wikipedia.org/wiki/George_E._P._Box).

Here is the approach using SQL commands.

```{r george_box1}
library(sqldf)
box <- dbConnect(SQLite(),
  dbname="../data/melange.sqlite")
george3 <- dbGetQuery(conn=box,  "
    pragma table_info(crawling)
")

george3
dbDisconnect(conn=box)
```

Here is an alternative approach using an R function. You don't have to do your homework both ways; I'm just showing different methods.

```{r george_box2}
library(sqldf)
box <- dbConnect(SQLite(),
  dbname="../data/melange.sqlite")
dbListFields(conn=box, name="crawling")
dbDisconnect(conn=box)
```
