---
title: "Homework14a"
author: "Suman Sahil and Steve Simon"
output:
  pdf_document: default
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, 
    encoding=encoding,
    output_dir = "../results", 
    output_format = "all") })  
---

This file was created on 2020-07-24 and last modified on `r Sys.Date()`.

Note: this solution uses R and SQLite. An alternate solution using SAS and Oracle is also available.

Note: Some of the names used in this code are arbitrary and you can choose whatever names you want. To emphasize which names can be modified at your discretion, I am using names of famous statisticians.

The statistician being honored in this code is [Helen Walker](https://en.wikipedia.org/wiki/Helen_M._Walker).

1. Compute the intersection of the ids from acupuncture_baseline_results and acupuncture_one_year_results. Display the first ten rows of data only.

```{r helen_walker1}
library(sqldf)
walker <- dbConnect(SQLite(),
  dbname="../data/melange.sqlite")
helen1 <- dbGetQuery(conn=walker,  "
    select id
      from acupuncture_baseline_results
    intersect
    select id
      from acupuncture_one_year_results
    limit 10
")

helen1
dbDisconnect(conn=walker)
```

2. Compute the union of the ids from acupuncture_baseline_results and acupuncture_one_year_results. Display the first ten rows of data only.

```{r helen_walker2}
library(sqldf)
walker <- dbConnect(SQLite(),
  dbname="../data/melange.sqlite")
helen2 <- dbGetQuery(conn=walker,  "
    select id
      from acupuncture_baseline_results
    union
    select id
      from acupuncture_one_year_results
    order by id
    limit 10
")

helen2
dbDisconnect(conn=walker)
```

3. In a previous module, you were asked to list the first ten ids that were in acupuncture_baseline_resuts but not in acupuncture_one_year_results. Use the set operator "minus" to achieve the same goal. Note: for SQLite, use "except" instead of "minus".

```{r helen_walker3}
library(sqldf)
walker <- dbConnect(SQLite(),
  dbname="../data/melange.sqlite")
helen3 <- dbGetQuery(conn=walker,  "
    select id
      from acupuncture_baseline_results
    except
    select id
      from acupuncture_one_year_results
    limit 10
")

helen3
dbDisconnect(conn=walker)
```

