---
title: "Overview of Module 2: Retrieving data using the SQL SELECT statement"
author: "Suman Sahil, Steve Simon"
output:
  html_document: default
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding,
  output_dir = "../results", output_format = "all") })  
---

This file was created on 2020-07-05 and last modified on `r Sys.Date()`.

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
suppressMessages(suppressWarnings(library(knitr)))
suppressMessages(suppressWarnings(library(magrittr)))
suppressMessages(suppressWarnings(library(sqldf)))
git <- "https://github.com/pmean/introduction-to-sql/blob/master/"
pan <- ""
```

### Objectives

+ In this module, you will learn  how to
  + Retrieve a limited number of fields
  + Retrieve distinct values
  + Change field names

### SQL keywords

+ Here are the SQL keywords introduced in this module:
  + as
  + from
  + select

### Data used in "Sorting and Restricting Data"

This page provides a bit more details about the database used in this module. Please note that there might be small inconsistencies between the Oracle version and the SQLite version of the data.

+ Airlines
  + [Data dictionary](`r path`/airlines-data-dictionary.txt)
  + Oracle: no download needed, use schema=melange
  + SQLite: download [melange.SQLite](`r path`/melange.SQLite)
  + There is a backup of this data in [CSV format](`r path`/airlines.csv). You should not need to access this backup.

+ Crawling
  + [Data dictionary](`r path`/crawling-data-dictionary.txt)
  + Oracle: no download needed, use schema=melange
  + SQLite: download [melange.SQLite](`r path`/crawling.SQLite)
  + There is a backup of this data in [CSV format](`r path`/crawling.csv). You should not need to access this backup.

+ Hospital
  + [Data dictionary](`r path`/hospital-data-dictionary.txt)
  + Oracle: no download needed, use schema=ehr
  + SQLite: download [melange.SQLite](`r path`/ehr.SQLite)
  + There is a backup of this data in [CSV format](`r path`/cigarettes.csv). You should not need to access this backup.

+ Patient_type
  + [Data dictionary](`r path`/patient-type-data-dictionary.txt)
  + Oracle: no download needed, use schema=ehr
  + SQLite: download [melange.SQLite](`r path`/ehr.SQLite)
  + There is a backup of this data in [CSV format](`r path`/patient_type.csv). You should not need to access this backup.

### Videos and handouts

+ Video02a. Retriveing data using the SQL select statement (15 minutes)
  + Available in [Panopto format](https://umkc.hosted.panopto.com/Panopto/Pages/Viewer.aspx?id=4b2a0448-a776-46eb-a17f-aaa7010bfcbb)
  + Slides, available in [Powerpoint format](`r git`results/select.pptx)

```{r}
path <- "https://github.com/classes-medb/5508/blob/master/data"
```

### Discussion02a: One minute feedback

Graded: 3 points, one point deduction if submitted after deadline

Review the videos in this module and briefly answer the following questions

+ What was the most important thing that you learned?
+ What concept(s) do you still find confusing?
+ What would you like to learn next?

Please keep your answers brief. You are welcome to say something generic like "I found nothing confusing" or "I found everything confusing" or "I agree with the previous comment."

### Homework02a: Practice using the crawling database

Graded: 3 points, one point deduction if submitted after deadline

Use the database crawling_db. Refer to the page "Data used in this module" for a description of the data and where you can download it. Oracle users do not need to download anything.

1. Read all three fields and all records
2. Change Temperature to Temperature_F
3. Put your code and the output in a single PDF file

### Homework02b: Practice using the patient_type database

Graded: 3 points, one point deduction if submitted after deadline

Use patient_type table in EHR Datamart. Refer to the page "Data used in this module" for a description of the data and where you can download it. Oracle users do not need to download anything.

1. Read all fields and  all records
2. Change Pat_type_desc to Patient_Type_Desc
3. Put your code and the output in a single PDF file

### Quiz02

Graded: 2 points

+ What delimiter do you need when selecting more than one field from a table?
  + one or more blanks
  + a comma
  + a semicolon
  + no delimiters are needed

+ What keyword do you use to rename a field?
  + as
  + equals
  + name
  + rename

Place both the questions 