---
title: "Overview of Module 4: Sorting and restricting data"
author: "Suman Sahil, Steve Simon"
output:
  html_document: default
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding,
  output_dir = "../results", output_format = "all") })  
---

This file was created on 2020-12-16 and last modified on `r Sys.Date()`.

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
suppressMessages(suppressWarnings(library(knitr)))
suppressMessages(suppressWarnings(library(magrittr)))
suppressMessages(suppressWarnings(library(sqldf)))
git <- "https://github.com/pmean/introduction-to-sql/blob/master/"
pan <- ""
```

### Objectives

+ In this module, you will learn  how to
  + Limit the rows  that are retrieved by a query
  + Sort the rows that are retrieved by a query
  + Select a random sample of records
 
### Data used in "Sorting and Restricting Data"

```{r}
source("overview_paths.R")
```

This page provides a bit more details about the database used in this module. Please note that there might be small inconsistencies between the Oracle version and the SQLite version of the data.

+ Airlines (see module02)

+ Cigarettes
  + [Data dictionary](`r data`/cigarettes-data-dictionary.txt)
  + Oracle: no download needed, use schema=melange
  + SQLite: download [melange.SQLite](`r data`/melange.SQLite)
  + There is a backup of this data in [CSV format](`r data`/cigarettes.csv). You should not need to access this backup.

+ Encounter
  + [Data dictionary](`r data`/encounter-data-dictionary.txt)
  + Oracle: no download needed, use schema=ehr
  + SQLite: download [melange.SQLite](`r data`/ehr.SQLite)
  + There is a backup of this data in [CSV format](`r data`/encounter.csv). You should not need to access this backup.

+ Fat
  + [Data dictionary](`r data`/fat-data-dictionary.txt)
  + Oracle: no download needed, use schema=melange
  + SQLite: download [melange.SQLite](`r data`/melange.SQLite)
  + There is a backup of this data in [CSV format](`r data`/fat.csv). You should not need to access this backup.

+ Hospital (see module02)

+ Patient_type (see module02) 
  
### Readings

### Videos and handouts

+ Video04a.  ( minutes)
  + Available in [panopto format](`r pan`/Viewer.aspx?id=01e81458-a1f2-4851-bfa8-aaa70134a584)
  + Slides, available in [Powerpoint format](`r results`/slides-and-speaker-notes-13.pptx)
  

### Discussion03a: One minute feedback

Graded: 3 points, one point deduction if submitted after deadline

Review the videos in this module and briefly answer the following questions

+ What was the most important thing that you learned?
+ What concept(s) do you still find confusing?
+ What would you like to learn next?

Please keep your answers brief. You are welcome to say something generic like "I found nothing confusing" or "I found everything confusing" or "I agree with the previous comment."

### Discuss03b. Review this week's readings

Graded: 3 points, one point deduction if submitted late.

Review the reading provide in this module and provide a brief summary (3 or 4 sentences). Was there anything in this reading that surprised you.

### Quiz03a. Introduction to Structured Query Language
  
Graded: 3 points, one point deduction if submitted late.

Please create a single PDF file that includes both our questions and your answers to the following.

1. There are several important benefits of a relational database. Name one of them.

2. What company developed SQL in the early 1970s?

3. What is a key?
