---
title: "Simple SQL examples within R: select and where"
author: "Steve Simon"
date: "September 12, 2017"
output: html_document
---

We will work with the database fat.sqlite. If you do not have this database, you can download it from the github site or you can create it using the program, input-first-file.Rmd.

The dbListTables function will list the names of all the tables in your database.

```{r list-tables}
library(sqldf)
db <- dbConnect(SQLite(), dbname="../data/fat.sqlite")
dbListTables(conn=db)
```

There is also a nice function for reading an entire table into R, dbReadTable, ...

```{r dbReadTable}
fd <- dbReadTable(conn=db, name="fat")
head(fd)
```

..., but I would encourage you to access the entire table instead using the SQL statement "select *". The asterisk tells SQL to select every field in the table.

```{r select-star}
fd <- dbGetQuery(conn=db,
 "SELECT * FROM fat")
dim(fd)
head(fd)
```

```{r select-single-field}
fd <- dbGetQuery(conn=db,
 "SELECT case_number FROM fat")
dim(fd)
head(fd)
```

To select more than one field from a table, separate the field names with commas.

```{r select-multiple-fields}
fd <- dbGetQuery(conn=db,
 "SELECT case_number, age, bmi FROM fat")
dim(fd)
head(fd)
```

To select a limited number of rows, use the LIMIT command. Different databases will have different syntaxes for this. For details, see https://www.w3schools.com/sql/sql_top.asp.

```{r select-rows}
fd <- dbGetQuery(conn=db,
 "SELECT case_number, age, bmi FROM fat
  LIMIT 10")
dim(fd)
head(fd)
```

You can select specific cases using the WHERE statement.

```{r specific-cases}
fd <- dbGetQuery(conn=db,
 "SELECT case_number, age, bmi FROM fat
  WHERE case_number=3")
dim(fd)
head(fd)
```

You can select a range of case_numbers.

```{r range-of-cases}
fd <- dbGetQuery(conn=db,
 "SELECT case_number, age, bmi FROM fat
 WHERE case_number BETWEEN 110 and 113")
dim(fd)
head(fd)
```

Or multiple values

```{r multiple-values}
fd <- dbGetQuery(conn=db,
 "SELECT case_number, age, bmi FROM fat
  WHERE case_number IN (207, 209, 214, 218, 219)")
dim(fd)
head(fd)
```

You can select on any variable, not just the case_number

```{r select-senior-citizens}
fd <- dbGetQuery(conn=db,
 "SELECT case_number, age, bmi FROM fat
  WHERE age >= 65")
dim(fd)
head(fd)
```

You can combine selections using Boolean logic (AND, OR, NOT)

```{r select-obese-senior-citizens}
fd <- dbGetQuery(conn=db,
 "SELECT case_number, age, bmi FROM fat
  WHERE age >= 65 AND
    bmi >= 30")
dim(fd)
head(fd)
```

You can change the names of your fields with the AS statement.

```{r as-statement}
fd <- dbGetQuery(conn=db,
 "SELECT case_number AS id, age AS age_in_years, bmi FROM fat")
dim(fd)
head(fd)
```


```{r shut-down-gracefully}
dbDisconnect(conn=db)
```

```
