---
title: "Limiting records retrieved using the WHERE statement"
author: "Steve Simon"
date: "2019-08-09"
output: powerpoint_presentation
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
suppressMessages(suppressWarnings(library(knitr)))
suppressMessages(suppressWarnings(library(magrittr)))
suppressMessages(suppressWarnings(library(sqldf)))
```

### Description of the fat database
+ The fat database
  + Single table, fat
  + Two measures of percentage body fat
  + Compare to various body dimensions
+ You can find a description of this data set at
  + http://jse.amstat.org/datasets/body.txt
  
<div class="notes">

We will work with the database fat.sqlite. The name of the database is the same as the name of the only table in the database.

</div>

### Selecting a limited number of records (1/2)
+ SQL code (SQLite)
```
select * 
  from fat
  limit 5
```
+ SQL code (Oracle)
```
select * 
  from fat
  where rownum <= 5
```
+ Other databases
```
???
```

<div class="notes">

Selecting a limited number of records is one area where different databases use different approaches. SQLite uses a limit statement. Oracle uses a rownum variable in the where statement. Other databases use a keyword like top or first.

This is just one example of where the standards committees have not caught up with the database designers.

</div>

### Selecting a limited number of records (2/2)

```{r limit}
library(sqldf)
db <- dbConnect(SQLite(), 
  dbname="../data/fat.sqlite")
first_five_records <- dbGetQuery(conn=db,
  "select case_number, age, bmi  
    from fat
    limit 5")
first_five_records
```

<div class="notes">

Here's what the results looks like.

</div>

### Warning!!!
+ Order of records in a database is arbitrary
+ You may prefer a random sample

<div class="notes">

Now the order in which the records are stored in a database is somewhat arbitrary and there is no guarantee that you will get the exact same rows, especially if the database is being regularly updated.

It does help to pick the first few records to get a rough idea of what your data table looks like, but there is a serious risk here. Often in a database, the weird or unusual records either float to the top of a table or sink to the bottom. You have no guarantee that there is any resemblance between the first few records and the parts of the table beyond the first few records.

You may, at times, prefer to select a random sample of records. You'll see how to do this in a future lecture.

</div>

### Select a specific record (1/2)
```{r specific-cases}
sql_code <- 
"select case_number, age, bmi 
  from fat
  where case_number=3"
```
+ SQL code
```
`r sql_code`
```

<div class="notes">

The fat dataset has a primary key, case_number, that is a sequential number starting at 1 and going to the last record (252). This is a nice thing to have, but you may not have it in most databases. It makes it very easy to pick out a specific record like the third one.

</div>

### Select a specific record (2/2)
```{r}
dbGetQuery(conn=db, sql_code)
```

<div class="notes">

Here's what the results looks like.

</div>

### Select a range of records (1/2)
```{r ragne-of-cases}
sql_code <- 
"select case_number, age, bmi 
  from fat
  where case_number BETWEEN 110 and 113"
```
+ SQL code
```
`r sql_code`
```

<div class="notes">

You can select a range of values with the BETWEEN keyword.

</div>

### Select a range of records (2/2)
```{r}
dbGetQuery(conn=db, sql_code)
```

<div class="notes">

Here's what the results looks like.

</div>

### Select a mix of records (1/2)
```{r multiple-values}
sql_code <- 
"SELECT case_number, age, bmi 
  FROM fat
  WHERE case_number IN (207, 209, 214)"
```
+ SQL code
```
`r sql_code`
```

<div class="notes">

You can use the IN keyword to select from a non-contiguous list of values.

</div>

### Select a mix of records (2/2)
```{r}
dbGetQuery(conn=db, sql_code)
```

<div class="notes">

Here's what the results looks like.

</div>

### Select on other variables (1/2)

```{r select-senior-citizens}
sql_code <- 
"SELECT case_number, age, bmi 
  FROM fat
  WHERE age >= 72"
```
+ SQL code
```
`r sql_code`
```

<div class="notes">

You can select on any variable, not just the case_number

</div>

### Select on other variables (2/2)

```{r}
dbGetQuery(conn=db, sql_code)
```

<div class="notes">

Here's what the results looks like.

</div>

### Selections with Boolean logic (1/2)
```{r boolean}
sql_code <- 
"SELECT case_number, age, bmi 
  FROM fat
  WHERE
    age >= 72 AND
    bmi >= 30"
```
+ SQL code
```
`r sql_code`
```

<div class="notes">

You can use standard Boolean logic with keywords like AND, OR, and NOT

</div>

### Selections with Boolean logic (2/2)

```{r}
dbGetQuery(conn=db, sql_code)
```

<div class="notes">

Here's what the results looks like. Notice that there is only one record because the "and" requires both statements to be true before a record is selected.

</div>

```{r shut-down-gracefully}
dbDisconnect(conn=db)
```

### Selections involving character data
+ Re-introduce airlines data set
  + Airline is a variable width character data type.

<div class="notes">

The fat data set has all numeric data. Selection using character data has a few distinctions.

</div>

### Remember what the full table looks like

```{r select-star}
library(sqldf)
db <- dbConnect(SQLite(), 
  dbname="../data/airlines_db.sqlite")
airlines_data <- dbGetQuery(conn=db,
 "select * 
    from airlines_table")
airlines_data
```

<div class="notes">

This is a table small enough that you can view everything on a single slide.

</div>

### Selecting a specific string (1/2)

```{r specific-string}
sql_code <- 
"select Airline, r2016, r2017
  from airlines_table
  where Airline >= 'UNITED'"
```
+ SQL code
```
`r sql_code`
```

<div class="notes">

Selecting a specific string is pretty much the same. You can use single quotes or double quotes, but you can not use the so-called "Smart" quotes that slant differently at the beginning than at the end.

</div>

### Selecting a specific string (2/2)
```{r}
dbGetQuery(conn=db, sql_code)
```

<div class="notes">

Here's what the results looks like. 

</div>

### Selecting a range of strings (1/2)

```{r range-of-strings}
sql_code <- 
"select Airline, r2016, r2017
  from airlines_table
  where Airline between 'U' and 'Z'"
```
+ SQL code
```
`r sql_code`
```

<div class="notes">

Selecting a specific string is pretty much the same. You can use single quotes or double quotes, but you can not use the so-called "Smart" quotes that slant differently at the beginning than at the end.

</div>

### Selecting a range of strings (2/2)
```{r}
dbGetQuery(conn=db, sql_code)
```

<div class="notes">

Here's what the results looks like. 

</div>

### Warning
+ What is between may not always be intuitive.
  + Is "w" between "U" and "Z"?
  + Where do numbers fit in?
  + How do special codes fit in (e.g, tabs versus spaces)?
  + How do accented characters fit in?
  
<div class="notes">

Ranges can be tricky. If you use this approach, run a few tests first so you know what you are getting. Do lower case letters fit in right next to their upper case counterparts? Where do the numbers fit in? There are certain things like tabs that can foul you up because they look just like spaces, but they have a different location in the order of things. If you have accented characters, which can occur easily for foreign names, make sure you know where they fit in as well. This is something that might depend on what country your computer resides in. European computers might produce different results than American computers for some of these queries.

</div>

### Selecting based on part of a string (1/6)

```{r like-front}
sql_code <- 
"select Airline, r2016, r2017
  from airlines_table
  where Airline like 'JET%'"
```
+ SQL code
```
`r sql_code`
```

<div class="notes">

If you are searching for just part of a string, use the keyword LIKE and include a percent sign for the portion of the string that you can leave open ended. The code shown above looks for an airline that has the letters J-E-T followed by anything (or nothing).

</div>

### Selecting based on part of a string (2/6)
```{r}
dbGetQuery(conn=db, sql_code)
```

<div class="notes">

Here's what the results looks like. 

</div>

### Selecting based on part of a string (3/6)

```{r like-back}
sql_code <- 
"select Airline, r2016, r2017
  from airlines_table
  where Airline like '%JET'"
```
+ SQL code
```
`r sql_code`
```

<div class="notes">

The code here looks similar, but %JET means that you can start with anything, but you have to end with J-E-T

</div>

### Selecting based on part of a string (4/6)
```{r}
dbGetQuery(conn=db, sql_code)
```

<div class="notes">

Here's what the results looks like. 

</div>

### Selecting based on part of a string (5/6)

```{r like-middle}
sql_code <- 
"select Airline, r2016, r2017
  from airlines_table
  where Airline like '%A%'"
```
+ SQL code
```
`r sql_code`
```

<div class="notes">

Here's the code that looks for the single letter A either in the middle or in the beginning or the end.

</div>

### Selecting based on part of a string (6/6)
```{r}
dbGetQuery(conn=db, sql_code)
```

<div class="notes">

Here's what the results looks like. 

</div>

### Selecting using NOT (1/2)

```{r not-like}
sql_code <- 
"select Airline, r2016, r2017
  from airlines_table
  where Airline not like '%A%'"
```
+ SQL code
```
`r sql_code`
```

<div class="notes">

The NOT keyword is useful for finding things that fail a particular match.

</div>

### Selecting using NOT (2/2)
```{r}
dbGetQuery(conn=db, sql_code)
```

<div class="notes">

Here's what the results looks like. 

</div>

### Your homework
+ The cigarettes database
  + Single table, cigarettes
  + Tar, nicotine, and carbon monoxide levels (mg)
  + Weight of cigarette in grams
+ You can find a description of this data set at
  + http://jse.amstat.org/datasets/cigarettes.txt
+ Find the records for cigarettes that have an ampersand (&) in their name
+ Find the records for cigarettes that are a single word brand name (no spaces)
+ Put your code and output in a single pdf file 

<div class="notes">

You will have a different data set for your homework. This data set measures all the bad things that you can find in a cigarette: tar, nicotine, and carbon monoxide. Find and print the records of any cigarettes whose brand name includes an ampersand. Also, find and print cigarettes whose brand names include just a single word. You'll need the NOT keyword here. Put all your results in a single PDF file.

</div>

